Product graphics

### 2009-11-26 product-graphics.tex
More work on paper

\title{Product graphics}
\author{Heike Hofmann and Hadley Wickham}
\begin{document}
\maketitle

@@

% JASA methods
%

% The Grammar of Graphics [reference] describes the independent components of
% statistical graphics and how they can be combined together. These components
% are very flexible and can be used to describe any type of plot. However, the
% book (Grammar with a capital G) does not present many examples of graphics
% for categorical data, and in those examples the use of the grammar gives
% little insight into plot. This paper describes common categorical plots, and
% presents a specialisation of the Grammar of Graphics specifically designed
% to describe these types of plots and give insight into the types of
% relationships that they can show.

Area plots [reference] display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination. The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots, equal-bin-size plots, and fluctuation diagrams. Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data. Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin. Figure \ref{fig:cat-examples} illustrates some of these plots.

@@

\begin{itemize}
  \item Put all area plots (bar charts, spine plots, mosaic plots, fluctuation diagrams) on a common footing
  \item Continue to build link between graphics and log-linear models
  \item Provide tools for ``residuals and re-iteration''
\end{itemize}

@@

Area plots are the graphical equivalent of contingency tables. Like contingency tables, they can be used to display raw counts, or conditioned and marginal breakdowns. Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.  We can decompose these graphics in a similar way to the way we decompose products of probabilities.

Let $f(x, y, z)$ be the 2d pmf function generated.  Can write this pmf as product of marginals and conditionals

@@

Generalised treemaps: \citep{vliegen:2006}
Cocktailmaps: \citep{ahlberg:1996}
Mosaic plots: \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
Extension to mosaic: \citep{friendly:1999,hofmann:2000}
Connection to models: \citep{theus:1999,hofmann:2001}

% Hierarchical coordinate systems 
%
% What are the common components of all these plots, and how can be describe
% them in a succinct way that also allows us to describe what tasks each
% display is good for?  We propose that there is one key feature that makes
% these plots special.  That is there their coordinate system, which is
% hierarchical.  
%
% In most coordinate systems that we are used to, such as Cartesian or polar
% coordinates, are symmetric in the sense that you can locate a point by
% starting from either dimension.  For example, in Cartesian coordinates you
% can identify either the x- or y-coordinate first, and the other next.  This
% also implies a distance symmetry when projected onto 2D Euclidean geometry.
% When you vary one coordinate by a small amount (holding all others constant)
% points in the projective geometry will be close.
%
% Neither of these properties hold for hierarchical coordinate systems. 
%
% We will discuss hierachical coordinate systems in general, and then present
% a specific hierachical coordinate system which gives rise to the area plots
% described previously.  We also present an R implementation, in the R package
% {\tt ggplot}.

To illustrate the techniques in this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from a yearly survey from 1972 to 2006.  Happy, age, sex, marital status, terminal education level, relative financial status and health.  Also contains a weighting variable wtssall, which is a probability weight from the survey.

>>>>

\title{Product graphics}
\author{Hadley Wickham and Heike Hofmann}
\begin{document}
\maketitle

@@

% JASA methods
% Infovis?

% Aims of this paper:
%
%  * emphasise the connection between area products and factorising 
%    probabilities
% 
%  * pull all area plots on a common footing - provide a grammar for 
%    categorical graphics

Area plots are the graphical equivalent of contingency tables, displaying how a total is broken down into components.

display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination.

The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003},  equal-bin-size plots, and fluctuation diagrams. Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data. Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.

Figure \ref{fig:cat-examples} illustrates some of these plots.

@@

Hierarchical partitioning of space.

@@

Let $f(x, y, z)$ be the 3d pmf function.  We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.

@@

% Other papers to cite:
% 
% Generalised treemaps: \citep{vliegen:2006}
% Cocktailmaps: \citep{ahlberg:1996}
% Extension to mosaic: \citep{friendly:1999,hofmann:2000}
% Connection to models: \citep{theus:1999,hofmann:2001}

% Hierarchical coordinate systems 
%
% What are the common components of all these plots, and how can be describe
% them in a succinct way that also allows us to describe what tasks each
% display is good for?  We propose that there is one key feature that makes
% these plots special.  That is there their coordinate system, which is
% hierarchical.  
%
% In most coordinate systems that we are used to, such as Cartesian or polar
% coordinates, are symmetric in the sense that you can locate a point by
% starting from either dimension.  For example, in Cartesian coordinates you
% can identify either the x- or y-coordinate first, and the other next.  This
% also implies a distance symmetry when projected onto 2D Euclidean geometry.
% When you vary one coordinate by a small amount (holding all others constant)
% points in the projective geometry will be close.
%
% Neither of these properties hold for hierarchical coordinate systems. 
%
% We will discuss hierachical coordinate systems in general, and then present
% a specific hierachical coordinate system which gives rise to the area plots
% described previously.

To illustrate these ideas, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.

\begin{itemize}
  \item Happy.  Discrete: ``very happy'', ``pretty happy'', ``not too happy''
  \item Age (in years), 18--89.
  \item Sex: ``female'', ``male''
  \item Degree: ``lt high school'', ``high school'', ``junior college'', ``bachelor'', ``gradudate''
  \item Relative financial status {\sf finrela}: ``far above average'', ``above average'', ``average'', ``below average'', ``far below average''
  \item Health: ``excellent'', ``good'', ``fair'', ``poor''
  \item Year: 1972--2006.
  \item Probability weight ({\sf wtsall}): 0.43--6.42.
\end{itemize}
